## Functions and Conditionals

### Conditionals

The basic syntax for an if statement goes like this:

---
---
title: "functions_loops_conditionals"
output: html_notebook
---

```{r}
if (logical statement) {
  do something
} else {
  do something else
}
#This is the idea 
```

The logical statements should boil down to a boolean; for example:

```{r}
a <- 10
if(a <= 5) {
  print(a)
} else {
  print('It is more than 5')
}
```

We can also add else if statements in the middle if we need more options.

```{r}
a <- 15
if(a <= 5) {
  print('Small')
} else if (a>5 & a<=10){
  print('Medium')
}else if (a> 10 | a == 7){
  print('Large')
} else {
  print('Huge!')
}
```

Neat! We can also nest conditionals just like we do in Python.

```{r}
a <- 5
if (a/2 != 0){
  if (a ==5){
    print('a is 5')
  }
}
```

Neat! Now, let's look at another core part of R programming - Loops!

### Loops

```{r}
for (value in 1:5){#for value in range 1 to 5
  #paste is a concatemating func
  print(paste(value, 'is greater than', value -1))

}
```

We can also do nifty stuff with our vectors, such as apply changes to specific indices of our data.

```{r}
a_vector <-  c(1,1,1,1,1,1)
#change the first third and sixth element to the value 6
for (i in c(1,3,6)){
  a_vector[i] <- 6
}
a_vector
```

We can loop over more than just lists:

```{r}
mat <-  matrix(1, nrow = 2, ncol = 2 )
mat
for (row in 1:2){
  for (col in 1:2){
    mat[col, row] <- 0 #order is important.. determines which direction the changes go
    print(mat)
  }
}

# nb: eg.. mat[col, row] <- 0 mat[row, col] <- 0
```

Like Python, we have while loops in R:

```{r}
b <-  0
while (b < 10){
  #increase b by 2
  b <- b + 2
  #print b
  print(b)
}
```

## Functions

When should we use functions?

```{r}
our_function <- function(optional_params){
  # do some stuff
}
our_function( pass something here)
```

An example:

```{r}
# a vector of temperatures
test_temp <- c(16.55055, 26.57600, 32.430220, 8.8392001, 20.42930)
  
  
  #func that converts from celcius to fahrenheit:
  c_to_f <- function(dat){
    temp_fahr <- dat*(9/5) + 32
    return(temp_fahr)
  }
  
print(c_to_f(26))
```

This is all cool.. but it didn't do anything. We need to call the function!

```{r}
data_fahr <-  c_to_f(test_temp)
data_fahr
```

Notice that in the background here, we are looping over each item in the array, but we never write a for loop! Keep this in mind, since when you pass in a large data set, the changes made by functions aren't quite immediate - the computer does have to go over each item in your data set, and change them one at a time.

A shortcut of sorts is to use the `apply()` function, which operates very similarly to Python's `map()`. Note that `apply` syntax goes as follows: `apply(X, MARGIN, FUN)` where X is the data set to use, MARGIN is specification for rows or columns, and FUN is the function to apply.

```{r}
#to map on purpose use apply()
test_mat <- matrix(c(1,1,1,1,2,2), nrow <- 2)
test_mat
apply(test_mat, 1, sum) # will add rows horizontally
apply(test_mat, 2, sum) # will add rows vertically
# 1 is for rows, 2 is for columns 
```

It is common to use `mean`, `median`, `sum`, `min`, or `max` with `apply`, but we can also use our user-defined functions as well. Note that apply() only works with matrices.

```{r}
#only works on matrices so
#create a matrix,
test_temp_matrix <- matrix(c(16.55055, 26.57600, 32.430220, 8.8392001, 20.42930), nrow = 2)

apply(test_temp_matrix, 1, c_to_f)
```

### Homework

1.  Do 5 CodeWars problems in R. Share your solutions here:

    ```{r}
    # 1. fucntion that calculates the average of the numbers in a given list. if list is empty, return  0

    #SOLUTIOM
    # Pseudo: average = mean = sum of list / length oflist

    find_average <- function(vec){
      sum_vec <- sum(vec)
      len_vec <- length(vec)
      if (len_vec > 0) {
        return(sum_vec / len_vec)
      }else {
        return (0)
      }
    }

    #test func
    nums <- c(1,2,3,4,5)
    print(find_average(nums))
    ```

    ```{r}
    #2. calculate the two oldest ages.
    # func should take an array of numbers as its argument and return the two highest numbers within the array
    # results = [second oldest, oldest]
    #SOLUTION
    #func takes numeric vector as input
    #func returns a vector with the two highest nums in the format [second oldest, oldest]
    #Pseudo: sort the ages in descending order
    #get the oldest age using index to get [1:2] 
    #use rev() to reverse order of ages
    two_oldest_ages <- function(ages){
      sort_ages <- sort(ages, decreasing = TRUE) #sorting the ages in descending order
      #to get two oldest age
      oldest_ages <- sort_ages[1:2]
      #return the ages in the format [second oldest, oldest]
      return(rev(oldest_ages))
    }

    #test func
    ages <- c(22, 95, 6, 79, 12, 24, 96, 68)
    print(two_oldest_ages(ages))
    ```

    ```{r}
    #3. function to find the smallest integer
    # Initialize smallest with the first number in the vector
    # smallest_num <- arr[1]
    findSmallestInt <- function(arr){
      sort_arr <- sort(arr)
      smallest_num <- sort_arr[1]#if (i in arr)
      return (smallest_num)
    }

    nums <- c(34, 15, 88, 2)
    print(findSmallestInt(nums))
    ```

    ```{r}
    #4. #function to calculate the minimum number of breaks needed
    #input will always be non-integer
    #Pseudo: check input validity
    #to calc min breaks needed n * m to get total num of squares the - 1 to keep last square
    break_chocolate <- function (n, m) {
      if(n <= 0 || m <= 0){
        return(0)
      }
      breaks_num <- n * m - 1
      return(breaks_num)
    }

    #test func
    print(break_chocolate(4, 1))
    ```

    ```{r}
    #This code does not execute properly. Try to figure out why.
    # mul <- function(a, b) {
    #   a * c # try to figure out why it doesn't work!
    # }

    # the variable c does not exist as a parameter / argument in the function defined..

    #working code
    mul <- function(a, b) {
      a * b # try to figure out why it doesn't work!
    }
    #test func

    print(mul(4,5))
    ```

2.  Create a function that checks an input of type `int` and returns fizz if the number is a multiple of 3, buzz if the number is a multiple of 5, and fizzbuzz if the number is a multiple of 3 and 5.

    ```{r}
    #use range to gen vector of 2 , 20
    #define function
    # ;input integer will be [num]
    # check if num is a multiple of both 3 and 5.... return fizzbuzz else if multiple of 3 else if multiple of 5... 
    # SOLUNTION: FUNCTION
    fizzbuzz_func <- function(num){
      if (num %% 3 == 0 && 5 == 0){
        return('fizzbuzz')
      } else if (num %% 3 == 0){
        return ('fizz')
      } else if (num %% 5 == 0){
        return('buzz')
      }else {
        return(num)
      }
    }

    #CALL FUNCTION
    numbers <- 2:20
    numbers
    # apply function to each num in list
    result <- sapply(numbers, fizzbuzz_func)
    result
    ```

3.  **Mario Challenge:** In the classic Super Mario games, Mario ends each level by ascending a flight of stairs and leaping off the top to arrive at the castle at the end. (See [here](https://youtu.be/-avspZlbOWU?t=55) for details). We can imagine that each of the bricks of the stairs is a character X. Create a function that prints out a set of stairs. For example:

    If the function is called marioStairs() and we pass in the value 4, we would expect to see stairs with a length of 4 and a height of 4:t

    X

    XX

    XXX

    XXXX

    ```{r}
    #if X is the character
    # Ascending order..
    #funct that prints out a set of stairs 
    #func loops from 1 to i. for each num_passed, func prints X , num_passed times
    #i will be each nbr in the sequence..
    #prints on a new line
    #func marioStairs()
    #func returns stairs with a length 4 and height 4
    #print X * (num_passed) eg X * 4what 
    #cat(): acts like print and concatenates without leaving a new line
    #replicate(): to replicate X, i nbr of times 
    marioStairs <- function(num_passed) {
      for (i in 1: num_passed){
        cat(replicate(i, 'X'), '\n') #every i instance on a new line 
      }
    } #func loops over nbrs, prints x on a new line for each i in the sequence

    #test func with 4
    marioStairs(4) 
    ```
